From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/discussion.tex
Abstract, some fixes

\section{Discussion \& Related Work}~\label{sec:discussion}

In this paper, we \ldots

\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

@@

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom. 
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesing a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua-\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
constructions are hard (or impossible in general case).

@@

\begin{center}
    \begin{tabular}{c c c}
        \hline
        Logic                              & Program                       & Algebra                     \\
        \hline
        Propositions                       & Types                         & Objects                     \\
        \hline
        Proofs                             & Terms                         & Morphisms                   \\
        \hline
        $\blank$                           & Equations                     & Homotopies                  \\
        \hline
        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
        \hline
        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
        \hline
        Connectives                        & Type constructors             & Bifunctors                  \\
        \hline
        ?                                  & Type normalization            & ?                           \\
        \hline
        ?                                  & Program normalization         & ?                           \\
        \hline
        ?                                  & Soundness of 2-combinators    & ?                           \\
    \end{tabular}
\end{center}

@@

>>>>

\section{Discussion \& Related Work}
\label{sec:discussion}

% In this paper, we \ldots

\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua_\beta}$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

@@

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom. 
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesing a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua_\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
constructions are hard (or impossible in the general case).

@@

\begin{center}
    \begin{tabular}{c c c}
        \hline
        Logic                              & Program                       & Algebra                     \\
        \hline
        Propositions                       & Types                         & Objects                     \\
        \hline
        Proofs                             & Terms                         & Morphisms                   \\
        \hline
        $\blank$                           & Equations                     & Homotopies                  \\
        \hline
        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
        \hline
        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
        \hline
        Connectives                        & Type constructors             & Bifunctors                  \\
        \hline
        ?                                  & Type normalisation            & ?                           \\
        \hline
        ?                                  & Program normalisation         & ?                           \\
        \hline
        ?                                  & Soundness of 2-combinators    & ?                           \\
    \end{tabular}
\end{center}

@@

\todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
